---
title: "Evaluate Temporal Overlap using Absolute Encounter Intensity"
author: "Yunyi Shen"
date: "10/27/2019"
output: pdf_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
require(dplyr)
require(ggplot2)
require(grid)
require(Rcpp)
source("R/Fourier_est.R")
sourceCpp("src/Fourier_calc.cpp")
vplayout <- function(x, y) viewport(layout.pos.row = x, layout.pos.col = y)

rawdata = read.csv("_data_/CleanEvents.csv",stringsAsFactors = F)
rawdata$Date_Time = as.POSIXct( paste(rawdata$Date,rawdata$Time),format = "%Y/%m/%d %H:%M:%OS")

rawdata = na.omit(rawdata)

mindate_st = min(rawdata$Date_Time[rawdata$Site=="Stockton"]) %>%
  as.character()%>%
  substr(1,10) %>%
  paste("00:00:00") %>%
  as.POSIXct(format = "%Y-%m-%d %H:%M:%OS") %>%
  as.numeric() # first second at the first detection day

rawdata$sec = as.numeric(rawdata$Date_Time)-as.numeric(mindate_st)# all in seconds

# species
spplist = unique(rawdata$Species)

# data of stockton
stockton = lapply(unique(rawdata$Species), function(spp,rawdata){
  rawdata[rawdata$Site=="Stockton" & rawdata$Species==spp,]
},rawdata)

names(stockton) = unique(rawdata$Species)

# time at event
event_time = lapply(stockton,function(kk){sort(kk$sec)})


spplist = names(event_time)[sapply(event_time, length)>3]

event_time_list = lapply(spplist, function(spp,event_time){
  event_time[[spp]]
},event_time)

names(event_time_list) = spplist

period = 24*3600 # period, a day, can be change based on sun

```

## Model Daily Activity Pattern using Time Inhomogeneous Poisson Process with Periodic Intensity Function

Detection of camera trapping can be viewed as time-to-event data. One common selection to model this type of process is Poisson point process. Two main assumptions for Poisson point process are:

- Independency: Number of events in any disjoint time intervals are independent

- Ordinariness:

$$
P(N(t+\delta t)-N(t)=1)=\lambda(t)\delta t + o(\delta t)
$$
$$
P(N(t+\delta t)-N(t)>1)=o(\delta t)
$$
From assumption 1, we can derive the waiting time ($T-T_0$, how long did it take from time $T_0$ to next event)'s distribution to be:
$$
P(T-T_0>t)=exp(\int_{T_0}^{T}\lambda(u)du)
$$

Denote $\int_{0}^{t}\lambda(u)du=\Lambda(t)$, then $\int_{T_0}^{T}\lambda(u)du=\Lambda(T)-\Lambda(T_0)$. The $\lambda(t)\ge 0$ is called *intensity function* and $\Lambda(t)$ is called *leading function* of the process. The *intensity function* can be understand as in a short time period, how many events (e.g. detections) should be expect.

From assumption 1 and 2, we derive the likelihood function for time-at-event. Denote time to the $ith$ event as $T_i$, then $X_i=T_i-T_{i-1}$ and $X_0=T_0$

$$
f_{T_1,T_2...T_n}=exp(-\Lambda(T_n))\prod_{i=0}^{n}\lambda(T_i)
$$


To model daily activity pattern from time-to-event data, we can consider a family of periodic intensity function with period of $P=1\ day=86400s$. To construct this family of positive periodic functions, we could use an exponential of Fourier series with amplitude parameters $A_i$ and phase parameters $\phi_i$ and $N$ terms, i.e. 

$$
\lambda (t)=exp(\frac{T_0}{2}+\sum_{i=1}^NA_icos(\frac{2\pi it}{P}-\phi_i))
$$

Then plug into likelihood function, we can have a MLE for parameters of the Fourier series and also an AIC to help us choose $N$.


## Result from the intensity function

From the intensity function of different species, we can derive different probabilities. Denote $\lambda_i(t)$ as the detection intensity of species $i$.

Then 
$$\int_{0}^P\lambda(u)du$$ 
is the detection probability for a day. 

Also we have:

$$
P(\text{it is spp i}|\text{detect spp i or j at t})=\frac{\lambda_i(t)}{\lambda_i(t)+\lambda_j(t)}
$$

## Some results with APIS, Stockton data:

AIC table for number of Fourier terms:

```{r echo = F,warning=FALSE,message=FALSE,error=FALSE, fig.height=6, fig.width=6}
set.seed(42)
MLE_fourier3 = lapply(event_time_list,function(ww){
  optim(runif(7),logL,n=3,event_time = ww,method = "BFGS",control = list(maxit = 1000))
}) # MLE using event-time

conv3 = sapply(MLE_fourier3,function(w){w$convergence}) # check convergence


AICs_3 = sapply(MLE_fourier3,function(ww){
  2*length(ww$par)+2*ww$value

})

MLE_fourier4 = lapply(event_time_list,function(ww){
  optim(runif(9),logL,n=4,event_time = ww,method = "BFGS",control = list(maxit = 1000))
}) # MLE using event-time

conv4 = sapply(MLE_fourier4,function(w){w$convergence}) # check convergence


AICs_4 = sapply(MLE_fourier4,function(ww){
  2*length(ww$par)+2*ww$value

})

AIC_table = rbind(AICs_3,AICs_4)
row.names(AIC_table) = c("N=3","N=4")

knitr::kable(AIC_table[,c("Bear_black","Coyote","Fox_red")],caption = "Sample AIC table")

```


Absolute intensity of fox and coyote:

```{r echo = F,warning=FALSE,message=FALSE,error=FALSE, fig.height=3, fig.width=9}
predic_fourier = lapply(MLE_fourier3,function(res){
  lambda_Fourier_series(seq(0,86400,1),3 # number of terms
                        ,res$par[1] # intercept
                        ,res$par[2:4] # amplitude
                        ,86400 # period, a day
                        ,res$par[5:7]) # phase
}) # predict one day's intensity function using the fitted Fourier series

predic_fourier4 = lapply(MLE_fourier4,function(res){
  lambda_Fourier_series(seq(0,86400,1),4 # number of terms
                        ,res$par[1] # intercept
                        ,res$par[2:5] # amplitude
                        ,86400 # period, a day
                        ,res$par[6:9]) # phase
})

detection_rate = lapply(MLE_fourier3,function(res){
  1-exp(-Lambda_Numeric_int(86400,3
                        ,res$par[1]
                        ,res$par[2:4]
                        ,86400
                        ,res$par[5:7],10000))
}) # detection rate per day calculated using integral of detection intensity for a day


times = seq(0,86400,1)

plotdata = data.frame(time = times,lambda = predic_fourier$Coyote,spp = "Coyote")
plotdata = rbind(plotdata,data.frame(time = times,lambda = predic_fourier4$Fox_red,spp = "Red Fox"),data.frame(time = times,lambda = predic_fourier$Bear_black,spp = "Bear"))

ratiodata = data.frame(time = times,ratio = predic_fourier$Coyote/(predic_fourier$Coyote+predic_fourier4$Fox_red),spp = "coyote-fox")

ratiodata = rbind(ratiodata,data.frame(time = times,ratio = predic_fourier$Bear_black/(predic_fourier$Bear_black+predic_fourier4$Fox_red),spp = "bear-fox"),data.frame(time = times,ratio = predic_fourier$Bear/(predic_fourier$Coyote+predic_fourier$Bear_black),spp = "bear-coyote"))

sepr = ggplot2::ggplot(plotdata,aes(x=time,y=lambda)) + 
  geom_line(aes(color = spp)) + 
  labs(x="time/s")+
  ggtitle("intensity")
  
ratio = ggplot(ratiodata,aes(x=time,y=ratio)) + 
  geom_line(aes(color=spp))+
  labs(x="time/s",y="lambda_c/(lambda_c+lambda_f)")+
  ggtitle("intensity ratio")

grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
print(sepr, vp = vplayout(1, 1))
print(ratio, vp = vplayout(1, 2))



Dettable = (unlist(detection_rate))



knitr::kable(Dettable[c("Bear_black","Coyote","Fox_red")],caption = "Daily Detection Probability",col.names = "P")

```

For fox, the intensity did not changed so much compare with coyote, and in most of the time, we should expect to see a coyote rather than a red fox.
